---
title: "Cytokine Resistance Phosphodata"
author: "Sara Gosline"
date: "4/9/2021"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
##process cytokine data
library(amlresistancenetworks)
library(dplyr)
```

## Phospho data kinase enrichment

For this analysis will load in the phosphoproteomic data collected in the Molm13 cell lines. We use the 
stoichiometry corrected proteomic data to focus explicitly on the changes in kinase activity. 

```{r data load,echo=FALSE}

###load all the data

phosData<-querySynapseTable('syn22986341')%>%subset(!is.nan(LogRatio))%>%
  mutate(Gene=unlist(Gene))%>%
  mutate(site=unlist(site))


clinvars<-phosData%>%
  dplyr::select(Sample='sample',CellType,TimePoint,Treatment)%>%
  distinct()

##what are we doing again?
summary<-phosData%>%
  dplyr::select(sample,CellType,TimePoint,Treatment)%>%
  distinct()%>%
  mutate(conditionName=stringr::str_c(CellType,TimePoint,Treatment,sep='_'))

print(summary)

phosMat<-phosData%>%dplyr::select(sample,site,LogRatio)%>%
  tidyr::pivot_wider(values_from=LogRatio,names_from=sample,
                     values_fn=list(LogRatio=mean,na.rm=T))%>%#,values_fill=list(LogRatio=0.0))%>%
  tibble::remove_rownames()%>%
  tibble::column_to_rownames('site')


```

## Now we can define the processing functions needed

We can plot the data and the kinase activity
```{r processing functions,echo=FALSE}
##
#' @param dat.table
plotAllData<-function(dat.table){
  library(ggfortify)
  met<-dat.table%>%dplyr::select(sample,CellType,TimePoint,Treatment)%>%
    distinct()
    
  mat<-dat.table%>%dplyr::select(Gene,LogRatio,sample)%>%
    distinct()%>%
    mutate(LogRatio=as.numeric(LogRatio))%>%
    tidyr::pivot_wider(names_from='sample',values_from='LogRatio',values_fn=list(LogRatio=function(x) mean(x,na.rm=T)),values_fill=list(LogRatio=0))%>%
  tibble::remove_rownames()%>%
    tibble::column_to_rownames('Gene')
  
  autoplot(prcomp(t(mat)),data=met,colour='Treatment',shape='CellType')
 
}

##plot kinase activity
plotKinDat<-function(kindat,sig.kin=NULL,prefix='all'){
  library(pheatmap)
  
  ##create matrix of kinase scores
  if(!is.null(sig.kin)){
    kindat<-subset(kindat,Kinase%in%sig.kin$Kinase.Gene)
    kinmat<-sig.kin%>%mutate(score='Yes')%>%
      tidyr::pivot_wider(names_from=Condition,values_from=score,values_fill=list(score='No'))%>%
      tibble::column_to_rownames('Kinase.Gene')
  
  }
  
  mat <-kindat%>%
    ungroup()%>%
    tidyr::pivot_wider(-c(meanNKINscore,numSubstr),
                                              values_from=meanLFC,
                                                names_from=Sample,
                                                values_fn=list(meanLFC=mean))%>%
  #,values_fill=0.0)%>%
    tibble::remove_rownames()%>%
    tibble::column_to_rownames('Kinase')
  
  if(!is.null(sig.kin))
    kinAts=kinmat
  else
    kinAts<-kindat%>%
      ungroup()%>%
      dplyr::select(Kinase,numSubstr)%>%
      distinct()%>%
      group_by(Kinase)%>%
      summarize(substrates=mean(numSubstr))%>%
      tibble::remove_rownames()%>%
      tibble::column_to_rownames('Kinase')
  
  sampAts<-phosData%>%
    dplyr::select(sample,TimePoint,Treatment,CellType)%>%
    distinct()%>%
    tibble::remove_rownames()%>%
    tibble::column_to_rownames('sample')
  
  sampAts$TimePoint=as.factor(sampAts$TimePoint)
  
  #vars=names(sort(apply(mat,1,var,na.rm=T),decreasing=T))
  pheatmap(mat,cellwidth = 8,cellheight=8,clustering_distance_cols = 'correlation',
          clustering_distance_rows = 'correlation',
          annotation_row = kinAts,annotation_col=sampAts) 

  pheatmap(mat,cellwidth = 8,cellheight=8,clustering_distance_cols = 'correlation',
          clustering_distance_rows = 'correlation',
          annotation_row = kinAts,annotation_col=sampAts,
          file=paste0(prefix,'cytokineKinaseHeatmap.pdf'),height=14,width=8) 
}

####Show kinase activity
kindat<-mapPhosphoToKinase(dplyr::rename(phosData,Sample='sample', LogFoldChange='LogRatio'))

```

## KSEA enrichment analysis
```{r ksea, echo=FALSE}
#' plot all the KSEA 
#' @param condList
#' @return data frame
doAllKSEAplots<-function(condList,pdat=phosData){
  
  gene.to.site<-dplyr::select(pdat,Gene,site,Peptide)%>%distinct()%>%
    dplyr::mutate(residue=stringr::str_replace(site,paste0(Gene,'-'),''))%>%
    dplyr::mutate(residue=stringr::str_replace_all(residue,"([STY])", ";\\1"))%>%
    dplyr::mutate(residue=stringr::str_replace(residue,"^;", ""))%>%
    dplyr::mutate(residue=stringr::str_replace_all(residue,"([sty])", ""))
  
  full.df<-purrr::map_df(names(condList),.f=function(clName){ 
    print(clName)
    condList[[clName]]%>%
      tibble::rownames_to_column('site')%>%
      left_join(gene.to.site)%>%
      dplyr::select(Gene,Peptide,residue,value='logFC',p_adj='adj.P.Val')%>%
      amlresistancenetworks::computeKSEA(.,prefix=clName,0.05)%>%
      mutate(Condition=clName)%>%
      as.data.frame()
  })
  return(full.df)
  
}


#' build networks from data frame
#' @param data.res
#' @param gene.col
#' @param weight.col
#' @param condition.col
#' @return network list?
runNetworksFromDF<-function(data,gene.col='Kinase.Gene',
                              weight.col='aveSubstrateLog2FC',
                              pvalue.col='P.Value',
                              condition.col='Condition',extra.col=c('Substrate.Gene','Source','log2FC'),
                              signif=0.01,beta=5){
  conditions=unique(dplyr::select(data,condition.col))
  for(con in conditions$condition){
    dat = data%>%
      dplyr::select(cond=condition.col,value=weight.col,Gene=gene.col,p.value=pvalue.col)%>%
      subset(p.value<signif)%>%
      subset(cond==con)
    vals<-dat$value
    names(vals)<- stringr::str_replace_all(dat$Gene,"([sty])", "")
    print(length(vals))
    #print(con)
    #print(vals)
    amlresistancenetworks::computePhosphoNetwork(phos.vals=vals,beta=beta,fname=paste0('phosphoNetworkfor',con,'_beta',beta))
  }
   
}
```

## Early time course comparisons

We want to identify changes in each condition at 5 min and 60 min compare to zero.

```{r early time course,warning=FALSE,echo=FALSE}

#phosMat<-fullPhosMat
#phosData<-uncorrectedPhosData
m13Phos<-list(Tram_5min=limmaTwoFactorDEAnalysis(phosMat,    
                        filter(summary,conditionName=='MOLM-13_5_Trametinib')$sample,
                         filter(summary,conditionName=='MOLM-13_0_none')$sample),
            Tram_60min=limmaTwoFactorDEAnalysis(phosMat,
                        filter(summary,conditionName=='MOLM-13_60_Trametinib')$sample,
                         filter(summary,conditionName=='MOLM-13_0_none')$sample),
              MCP1_5min=limmaTwoFactorDEAnalysis(phosMat,
                         filter(summary,conditionName=='MOLM-13_5_MCP-1')$sample,               
                         filter(summary,conditionName=='MOLM-13_0_none')$sample),
                MCP1_60min=limmaTwoFactorDEAnalysis(phosMat,                                    
                        filter(summary,conditionName=='MOLM-13_60_MCP-1')$sample,
                        filter(summary,conditionName=='MOLM-13_0_none')$sample),
                MCP1_tram_5min=limmaTwoFactorDEAnalysis(phosMat,
                        filter(summary,conditionName=='MOLM-13_5_Trametinib+MCP-1')$sample,
                        filter(summary,conditionName=='MOLM-13_0_none')$sample),
                MCP1_tram_60min=limmaTwoFactorDEAnalysis(phosMat,     
                        filter(summary,conditionName=='MOLM-13_60_Trametinib+MCP-1')$sample,
                        filter(summary,conditionName=='MOLM-13_0_none')$sample))
 
 p3<-doAllKSEAplots(m13Phos,phosData)
  
  combPhos<-list(comb_5min=limmaTwoFactorDEAnalysis(phosMat,
                         filter(summary,conditionName%in%c('MOLM-13_5_Trametinib','MOLM-13_5_MCP-1'))$sample,
                          filter(summary,conditionName=='MOLM-13_5_Trametinib+MCP-1')$sample),
                 comb_60min=limmaTwoFactorDEAnalysis(phosMat,
                          filter(summary,conditionName%in%c('MOLM-13_60_Trametinib','MOLM-13_60_MCP-1'))$sample,
                          filter(summary,conditionName=='MOLM-13_60_Trametinib+MCP-1')$sample))
    p4<-doAllKSEAplots(combPhos,phosData)

  combDf<-do.call(rbind,lapply(names(combPhos),function(x) mutate(combPhos[[x]],condition=x)))
  
  #tramMCP=runNetworksFromDF(combDf,gene.col='featureID',
  #                            weight.col='logFC',pvalue.col='P.Value',beta=5,
  #                            condition.col='condition',extra.col=c('AveExpr','t','P.Value'))
  #  tramMCP=runNetworksFromDF(combDf,gene.col='featureID',
  #                            weight.col='logFC',pvalue.col='P.Value',beta=10,
  #                            condition.col='condition',extra.col=c('AveExpr','t','P.Value'))
  #lateHeatmap<-kseaZscoreHeatmap(list(p3,p5),'earlyTreatmentKSEAzscoreHeatmap.pdf')
  sig.kin<-p4%>%
    subset(p.value<0.05)%>%
    select(Kinase.Gene,Condition)%>%distinct()
 # print(sig.kin)
  
  plotKinDat(subset(kindat,Sample%in%subset(summary,CellType=='MOLM-13')$sample),
             sig.kin,'timeCourse')

```

## Late time course comparisons
We want to see how the late resistant cells compared

```{r collect comparisons,echo=FALSE}
#####now do various comparisons

latePhos<-list(lateTram_vs_lateCombo=limmaTwoFactorDEAnalysis(phosMat,
                        filter(summary,conditionName=='Late MOLM-13_0_Trametinib+MCP-1')$sample,
                        filter(summary,conditionName=='Late MOLM-13_0_Trametinib')$sample),
               resist_vs_mcp1_5min=limmaTwoFactorDEAnalysis(phosMat,
                        filter(summary,conditionName=='MOLM-13 Tr Resistant_5_MCP-1')$sample,
                        filter(summary,conditionName=='MOLM-13 Tr Resistant_0_none')$sample),
              resist_vs_mcp1_60min=limmaTwoFactorDEAnalysis(phosMat,
                        filter(summary,conditionName=='MOLM-13 Tr Resistant_60_MCP-1')$sample,
                        filter(summary,conditionName=='MOLM-13 Tr Resistant_0_none')$sample))
  
  ph3<-doAllKSEAplots(latePhos)
             
#  earlyLatePh<-plotConditionsInFlow(earlyLatePhos,title='Phosphoproteomics in late',0.05)
#  ggsave('earlyLatePhos.png',earlyLatePh,width=11,height=6)
  
  latePhresdf<-do.call(rbind,lapply(names(latePhos),function(x) data.frame(latePhos[[x]],Condition=x)))
   # lateNets<-runNetworksFromDF(ph4)
  
   #tramMCP=runNetworksFromDF(latePhresdf,gene.col='featureID',
  #                            weight.col='logFC',pvalue.col='adj.P.Val',beta=5,
  #                            condition.col='Condition',extra.col=c('AveExpr','t','P.Value'))
  #      tramMCP=runNetworksFromDF(latePhresdf,gene.col='featureID',
  #                            weight.col='logFC',pvalue.col='adj.P.Val',beta=10,
  #                            condition.col='Condition',extra.col=c('AveExpr','t','P.Value'))
  sig.kin<-ph3%>%
    subset(p.value<0.05)%>%
    select(Kinase.Gene,Condition)%>%distinct()
#  print(sig.kin)
  
  plotKinDat(subset(kindat,
                    Sample%in%subset(summary,CellType%in%c('MOLM-13 Tr Resistant','Late MOLM-13'))$sample),
             sig.kin,'lateResistance')

```
These represent the late resistance samples

## Network analysis
To do the network analysis we focus on the substrates only, from the un-corrected data.

```{r network, echo=FALSE}
uncorrectedPhosData<-querySynapseTable('syn24389738')%>%subset(!is.nan(LogRatio))%>%
  mutate(Gene=unlist(Gene))%>%
  mutate(site=unlist(site))

fullPhosMat<-uncorrectedPhosData%>%dplyr::select(sample,site,LogRatio)%>%
  tidyr::pivot_wider(values_from=LogRatio,names_from=sample,
                     values_fn=list(LogRatio=mean))%>%#,values_fill=list(LogRatio=0.0))%>%
  tibble::remove_rownames()%>%
  tibble::column_to_rownames('site')

 uncorrectedKinDat<-mapPhosphoToKinase(dplyr::rename(uncorrectedPhosData,Sample='sample', LogFoldChange='LogRatio'))

 combPhos<-list(comb_5min=limmaTwoFactorDEAnalysis(fullPhosMat,
                          filter(summary,conditionName=='MOLM-13_5_Trametinib+MCP-1')$sample,
                          filter(summary,conditionName%in%c('MOLM-13_5_Trametinib','MOLM-13_5_MCP-1'))$sample),
                 comb_60min=limmaTwoFactorDEAnalysis(fullPhosMat,
                          filter(summary,conditionName=='MOLM-13_60_Trametinib+MCP-1')$sample,
                          filter(summary,conditionName%in%c('MOLM-13_60_Trametinib','MOLM-13_60_MCP-1'))$sample))
##now we can load this into a network
```