---
title: "Cytokine Resistance Phosphodata Experiment 2"
author: "Sara Gosline"
date: "9/9/2021"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
##process cytokine data
library(amlresistancenetworks)
library(dplyr)
```

## Phospho data kinase enrichment

For this analysis will load in the phosphoproteomic data collected in the Molm13 cell lines. We use the stoichiometry corrected proteomic data to focus explicitly on the changes in kinase activity.  

```{r data load,echo=FALSE,warning=FALSE}

###load all the data

phosData<-querySynapseTable('syn22986341')%>%subset(!is.nan(LogRatio))%>%
  mutate(Gene=unlist(Gene))%>%
  mutate(site=unlist(site))


clinvars<-phosData%>%
  dplyr::select(Sample='sample',CellType,TimePoint,Treatment,Batch)%>%
  distinct()

##what are we doing again?
earlySummary<-phosData%>%
    subset(Batch=='Experiment 1')%>%
  dplyr::select(sample,CellType,TimePoint,Treatment,Batch)%>%
  distinct()%>%
  mutate(conditionName=stringr::str_c(CellType,TimePoint,Treatment,sep='_'))

lateSummary<-phosData%>%
    subset(Batch=='Experiment 2')%>%
  dplyr::select(sample,CellType,TimePoint,Treatment,Batch)%>%
  distinct()%>%
  mutate(conditionName=stringr::str_c(CellType,TimePoint,Treatment,sep='_'))%>%
  subset(TimePoint==0)


earlyPhosMat<-phosData%>%
  subset(Batch=='Experiment 1')%>%
  dplyr::select(sample,site,LogRatio)%>%
  tidyr::pivot_wider(values_from=LogRatio,names_from=sample,
                     values_fn=list(LogRatio=mean,na.rm=T))%>%#updated 9/6,values_fill=list(LogRatio=0.0))%>%
  tibble::remove_rownames()%>%
  tibble::column_to_rownames('site')



latePhosMat<-phosData%>%
  subset(TimePoint==0)%>%
  subset(Batch=='Experiment 2')%>%
  dplyr::select(sample,site,LogRatio)%>%
  tidyr::pivot_wider(values_from=LogRatio,names_from=sample,
                     values_fn=list(LogRatio=mean,na.rm=T))%>%#updated 9/6,values_fill=list(LogRatio=0.0))%>%
  tibble::remove_rownames()%>%
  tibble::column_to_rownames('site')
```

## Experiment 2 data


We can plot the data and the kinase activity as needed in heatmap or PCA form. Here we want to visualize only the samples measured in Experiment 2 (minute the 120 min time point)
```{r processing functions,echo=TRUE,warning=FALSE}
##
#' @param dat.table
plotAllData<-function(dat.table,prefix=''){
  library(ggfortify)
  met<-dat.table%>%dplyr::select(sample,CellType,TimePoint,Treatment)%>%
    distinct()
    
  mat<-dat.table%>%dplyr::select(Gene,LogRatio,sample)%>%
    distinct()%>%
    mutate(LogRatio=as.numeric(LogRatio))%>%
    tidyr::pivot_wider(names_from='sample',values_from='LogRatio',values_fn=list(LogRatio=function(x) mean(x,na.rm=T)),values_fill=list(LogRatio=0))%>%
  tibble::remove_rownames()%>%
    tibble::column_to_rownames('Gene')
  
  
  p<-autoplot(prcomp(t(mat)),data=met,colour='Treatment',shape='CellType')
  
  ggsave(file=paste0('pcaPlot',prefix,'.pdf'),p, device='pdf')
  return(p) 
}

##plot kinase activity
plotKinDat<-function(kindat,sig.kin=NULL,prefix='all'){
  library(pheatmap)
  
  ##create matrix of kinase scores
  if(!is.null(sig.kin)){
    kindat<-subset(kindat,Kinase%in%sig.kin$Kinase.Gene)
    kinmat<-sig.kin%>%mutate(score='Yes')%>%
      tidyr::pivot_wider(names_from=Condition,values_from=score,values_fill=list(score='No'))%>%
      tibble::column_to_rownames('Kinase.Gene')
  
  }
  
  mat <-kindat%>%
    ungroup()%>%
    tidyr::pivot_wider(-c(meanNKINscore,numSubstr),
                                              values_from=meanLFC,
                                                names_from=Sample,
                                                values_fn=list(meanLFC=mean), values_fill=0.0)%>%
    tibble::remove_rownames()%>%
    tibble::column_to_rownames('Kinase')
  
  if(!is.null(sig.kin))
    kinAts=kinmat
  else
    kinAts<-kindat%>%
      ungroup()%>%
      dplyr::select(Kinase,numSubstr)%>%
      distinct()%>%
      group_by(Kinase)%>%
      summarize(substrates=mean(numSubstr))%>%
      tibble::remove_rownames()%>%
      tibble::column_to_rownames('Kinase')
  
  sampAts<-phosData%>%
    dplyr::select(sample,TimePoint,Treatment,CellType)%>%
    distinct()%>%
    tibble::remove_rownames()%>%
    tibble::column_to_rownames('sample')
  
  sampAts$TimePoint=as.factor(sampAts$TimePoint)
  
  #vars=names(sort(apply(mat,1,var,na.rm=T),decreasing=T))
 
  pheatmap(mat,cellwidth = 8,cellheight=8,clustering_distance_cols = 'correlation',
          clustering_distance_rows = 'correlation',
          annotation_row = kinAts,annotation_col=sampAts,
          file=paste0(prefix,'cytokineKinaseHeatmap.pdf'),height=14,width=8) 
  
  res<- pheatmap(mat,cellwidth = 8,cellheight=8,clustering_distance_cols = 'correlation',
          clustering_distance_rows = 'correlation',
          annotation_row = kinAts,annotation_col=sampAts) 
  return(res)
}

####Show kinase activity
kindat<-mapPhosphoToKinase(dplyr::rename(phosData,Sample='sample', LogFoldChange='LogRatio'))

##plot PCA
phosData%>%
  subset(Batch=='Experiment 2')%>%
  subset(TimePoint==0)%>%
  plotAllData('Experiment 2')
```


## KSEA enrichment analysis

Next we define functions to carry out various KSEA analyses. 
```{r ksea, echo=FALSE}
#' plot all the KSEA 
#' @param condList
#' @return data frame
doAllKSEAplots<-function(condList,pdat=phosData){
  
  gene.to.site<-dplyr::select(pdat,Gene,site,Peptide)%>%distinct()%>%
    dplyr::mutate(residue=stringr::str_replace(site,paste0(Gene,'-'),''))%>%
    dplyr::mutate(residue=stringr::str_replace_all(residue,"([STY])", ";\\1"))%>%
    dplyr::mutate(residue=stringr::str_replace(residue,"^;", ""))%>%
    dplyr::mutate(residue=stringr::str_replace_all(residue,"([sty])", ""))
  
  full.df<-purrr::map_df(names(condList),.f=function(clName){ 
    print(clName)
    condList[[clName]]%>%
      tibble::rownames_to_column('site')%>%
      left_join(gene.to.site)%>%
      dplyr::select(Gene,Peptide,residue,value='logFC',p_adj='adj.P.Val')%>%
      amlresistancenetworks::computeKSEA(.,prefix=clName,0.1,linkedSubs=1,suffix='pdf')%>%
      mutate(Condition=clName)%>%
      as.data.frame()
  })
  return(full.df)
  
}


#' build networks from data frame
#' @param data.res
#' @param gene.col
#' @param weight.col
#' @param condition.col
#' @return network list?
runNetworksFromDF<-function(data,gene.col='Kinase.Gene',
                              weight.col='aveSubstrateLog2FC',
                              pvalue.col='P.Value',
                              condition.col='Condition',extra.col=c('Substrate.Gene','Source','log2FC'),
                              signif=0.01,beta=5){
  conditions=unique(dplyr::select(data,condition.col))
  for(con in conditions$condition){
    dat = data%>%
      dplyr::select(cond=condition.col,value=weight.col,Gene=gene.col,p.value=pvalue.col)%>%
      subset(p.value<signif)%>%
      subset(cond==con)
    vals<-dat$value
    names(vals)<- stringr::str_replace_all(dat$Gene,"([sty])", "")
    print(length(vals))
    #print(con)
    #print(vals)
    amlresistancenetworks::computePhosphoNetwork(phos.vals=vals,beta=beta,fname=paste0('phosphoNetworkfor',con,'_beta',beta))
  }
   
}
```

### Combination time course comparisons

We want to identify changes in each condition at 5 min and 60 min compare to zero.

```{r early time course,warning=FALSE,echo=FALSE}

phosMat <- latePhosMat
summary <- lateSummary
#phosMat<-fullPhosMat
#phosData<-uncorrectedPhosData
m13Phos<-list(Tram_vs_control=limmaTwoFactorDEAnalysis(phosMat,
                         filter(summary,conditionName=='MOLM-13_0_none')$sample,    
                        filter(summary,conditionName=='Late MOLM-13_0_Trametinib')$sample),
            TramMCP1_vs_control=limmaTwoFactorDEAnalysis(phosMat,
                         filter(summary,conditionName=='MOLM-13_0_none')$sample,
                        filter(summary,conditionName=='Late MOLM-13_0_Trametinib+MCP-1')$sample),
              TramMCP1_vs_Tram=limmaTwoFactorDEAnalysis(phosMat,
                         filter(summary,conditionName=='Late MOLM-13_0_Trametinib')$sample,             
                         filter(summary,conditionName=='Late MOLM-13_0_Trametinib+MCP-1')$sample))
 
  p3<-doAllKSEAplots(m13Phos,phosData)
  
  sig.kin<-p3%>%
    subset(p.value<0.1)%>%
    select(Kinase.Gene,Condition)%>%distinct()
 # print(sig.kin)
  
  res=plotKinDat(subset(kindat,Sample%in%summary$sample), sig.kin,'exp2Results')
  res  
```

And there we have the analysis