---
title: "Final Manuscript Figures"
author: "Sara Gosline"
date: "2023-06-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(amlresistancenetworks)
library(dplyr)
```

## Description of work

This script only plots the data that was used in the manuscript, thus reducing the confusion of other documents. 

## Load data
Here we load the data from files located on Synapse at http://synapse.org/ptrc 

It is primarily two files.
```{r load data}
sync<-synapseLogin()
phosData<-readr::read_csv(sync$get('syn51753813')$path)|>#querySynapseTable('syn22986341')%>%subset(!is.nan(LogRatio
  subset(site!='')|>
 # mutate(Gene=unlist(Gene))%>%
#  mutate(site=unlist(site))%>%
  subset(TimePoint!=120)

globalData<-readr::read_csv(sync$get('syn51753816')$path)|>#querySynapseTable('syn22986326')%>%
  subset(!is.nan(LogRatio))%>%
  subset(Gene!='')|>
  #mutate(Gene=unlist(Gene))%>%
  subset(TimePoint!=120)

summary <- phosData %>%
  dplyr::select(sample, CellType, TimePoint, Treatment,Batch) %>%
  distinct() %>%
  mutate(conditionName = stringr::str_c(CellType, TimePoint, Treatment,
                                        sep = '_'))

print(summary)
```
Now that we have the data we can analyze it. 

## Data is from two distinct experiments

Below are the summaries of data from each experiment

### Experiment 1, Molm-13
```{r}
#' @param dat.table
plotAllData <- function(dat.table, expand=0.01, alpha=0.1, ...) {
  library(ggfortify)
  met <- dat.table %>%
    dplyr::select(sample, CellType, TimePoint, Treatment) %>%
    distinct()
    
  mat <- dat.table %>% dplyr::select(Gene,LogRatio,sample) %>%
    distinct() %>%
    mutate(LogRatio=as.numeric(LogRatio)) %>%
    tidyr::pivot_wider(names_from='sample', values_from='LogRatio',
                       values_fn=list(LogRatio=function(x) mean(x,na.rm=T)),
                       values_fill=list(LogRatio=0)) %>%
    tibble::remove_rownames() %>%
    tibble::column_to_rownames('Gene')
  mat <- mat[complete.cases(mat),]
  x <- prcomp(t(mat))$x
  x <- as.data.frame(x)
  x$sample <- rownames(x)
  ggdata <- inner_join(x, met, by="sample")
  
  library(ggforce)
  ggplot(ggdata, aes_string(x="PC1", y="PC2", ...)) +
    geom_point(size=2.5) +
    ggforce::geom_mark_ellipse(color = NA, alpha=alpha, expand=expand)
 
}

#pdf(file.path(output_dir, "Late_MOLM13_w_Parental_phospho_PCA.pdf"))
globalData %>% filter(Batch == "Experiment 1",
                    CellType == "MOLM-13") %>%
    dplyr::mutate(TimePoint=as.factor(TimePoint))|>
  plotAllData(colour='Treatment', fill='Treatment',shape='TimePoint') +
  ggtitle("Global PCA, Experiment 1")
#dev.off()


#pdf(file.path(output_dir, "Late_MOLM13_w_Parental_phospho_PCA.pdf"))
phosData %>% filter(Batch == "Experiment 1",
                    CellType == "MOLM-13") %>%
  dplyr::mutate(TimePoint=as.factor(TimePoint))|>
  plotAllData(colour='Treatment', fill='Treatment',shape='TimePoint') +
  ggtitle("Phospho PCA,Experiment 1")
#dev.off()

```

### Experiment 2
Here we have late resistance compared to early resistance.
```{r}

x <- globalData %>% filter(Batch == "Experiment 2",
                      TimePoint != 120)
x %>% plotAllData(color='Treatment', fill='Treatment',shape='CellType') +
  ggtitle("Global PCA, Late MOLM-13 + Parental")

phosData %>% filter(Batch == "Experiment 2",
                    TimePoint != 120) %>%
  plotAllData(colour='Treatment', fill='Treatment', shape='CellType') +
  ggtitle("Phospho PCA, Late MOLM-13 + Parental")

```

## Figures

We have multiple panels for each figure.

### Figure 3B

For figure 3B we focus on experiment 1 and analyze the phosphoproteomic changes at three points. 

```{r plotting, echo=F, warning=F, message=F}

#' plot all the KSEA 
#' @param condList
#' @return data frame
#' @importsFrom dplyr %>% distinct mutate left_join select 
#' @importsFrom stringr str_replace str_replace_all
#' @importsFrom purrr map_df
#' @importsFrom tibble rownames_to_column
#' @importsFrom amlresistancenetworks computeKSEA

library(stringr)
library(dplyr)
library(tibble)
library(purrr)

doAllKSEAplots <- function(condList, pdat, output_dir, p_cutoff=0.05) {
  gene.to.site<-dplyr::select(pdat,Gene,site,Peptide)%>%distinct()%>%
    dplyr::mutate(residue=stringr::str_replace(site,paste0(Gene,'-'),''))%>%
    dplyr::mutate(residue=stringr::str_replace_all(residue,"([STY])", ";\\1"))%>%
    dplyr::mutate(residue=stringr::str_replace(residue,"^;", ""))%>%
    dplyr::mutate(residue=stringr::str_replace_all(residue,"([sty])", ""))
  
  full.df <- purrr::map_df(names(condList), .f = function(clName) { 
    condList[[clName]] %>%
      rownames_to_column('site') %>%
      left_join(gene.to.site) %>%
      select(Gene, Peptide, residue, value='logFC', p_adj='adj.P.Val') %>%
      amlresistancenetworks::computeKSEA(., prefix = clName, p_cutoff, suffix='pdf') %>%
      mutate(Condition = clName) %>%
      as.data.frame()
  })
  
  return(full.df)
}

compare_samples <- function(phospho_data, treatment_A, treatment_B) {
  A_samples <- phospho_data %>%
    filter(Treatment == treatment_A) %>% 
    distinct(sample) %>%
    pull(sample)
  B_samples <- phospho_data %>%
    filter(Treatment == treatment_B) %>% 
    distinct(sample) %>%
    pull(sample)
  phosMat <- phospho_data_to_matrix(phospho_data)
  limmaTwoFactorDEAnalysis(phosMat, A_samples, B_samples)
}


fmean<-function(x){
  mean(x,na.rm=T)
}
phospho_data_to_matrix <- function(phospho_data){
  phospho_data %>%
    dplyr::select(sample, site, LogRatio) %>%
    tidyr::pivot_wider(values_from = LogRatio, names_from = sample,
                       values_fn = list(LogRatio = fmean)) %>%
    tibble::remove_rownames() %>%
    tibble::column_to_rownames('site')
}
phosMat <- phospho_data_to_matrix(phosData)
```


```{r fig3b,message=F, echo=T}
x <- phosData %>%
  filter(Batch == "Experiment 1")|>
  subset(TimePoint!=5)

limma_DEA_results <- list(Tram_vs_Parental = compare_samples(x,"none","Trametinib"),
                TramMCP1_vs_Parental = compare_samples(x,"none","Trametinib+MCP-1"),
                Tram_vs_TrMCP1 = compare_samples(x,"Trametinib+MCP-1","Trametinib"),
               MCP1_vs_Parental = compare_samples(x,"none","MCP-1"))
KSEA_plots_output <- doAllKSEAplots(limma_DEA_results, x, output_dir = ".", p_cutoff=0.1)
#output results
write.table(KSEA_plots_output, file='exp1_KSEA.tsv',
            quote=FALSE, sep='\t', col.names = NA)
```
Three of the four of these figures are in the paper.

### Supplemental Figure 3A
For this figure we focus on the proteomic analysis and compare sensitive versus enrichment samples. This requires only experiment 2.

```{r resistance vs sens, echo=T}

#' plot all the GO 
#' @param condList
#' @return data frame
doAllGOplots<-function(condList){
  
  full.df<-purrr::map_df(names(condList),.f=function(clName){ 
    condList[[clName]]%>%
      tibble::rownames_to_column('Gene')%>%
      dplyr::select(Gene,value='logFC')%>%
      amlresistancenetworks::plotOldGSEA(.,prefix=clName,0.1,width=16)%>%
      as.data.frame()
  })
  
}
protSumm <-subset(summary,Batch=='Experiment 2')


protMat<-globalData%>%
  # subset(Batch=='Experiment 1')%>%
    subset(TimePoint==0)%>%
  dplyr::select(sample,Gene,LogRatio)%>%
  distinct()|>
  tidyr::pivot_wider(values_from=LogRatio,names_from=sample,
                     values_fn=list(LogRatio=fmean))%>%
  as.data.frame()|>
  subset(Gene!='')|>
  tibble::remove_rownames()|>
  tibble::column_to_rownames('Gene')

m13Prot<-list(Tram_vs_control=limmaTwoFactorDEAnalysis(protMat,
                         filter(protSumm,conditionName=='MOLM-13_0_none')$sample,    
                        filter(protSumm,conditionName=='Late MOLM-13_0_Trametinib')$sample),
            TramMCP1_vs_control=limmaTwoFactorDEAnalysis(protMat,
                         filter(protSumm,conditionName=='MOLM-13_0_none')$sample,
                        filter(protSumm,conditionName=='Late MOLM-13_0_Trametinib+MCP-1')$sample),
              TramMCP1_vs_Tram=limmaTwoFactorDEAnalysis(protMat,
                         filter(protSumm,conditionName=='Late MOLM-13_0_Trametinib')$sample,      
                         filter(protSumm,conditionName=='Late MOLM-13_0_Trametinib+MCP-1')$sample))
 
  p3<-doAllGOplots(m13Prot)
write.table(p3, file='exp2_GSEA.tsv',
            quote=FALSE, sep='\t', col.names = NA)
```

The last figure to generate is the heatmap.?

