---
title: "DIA targeted differential expression & GSEA"
author: "Camilo Posso"
date: "09/13/2021"
output: 
  html_document:
    code_folding: hide
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

```{r libraries, include=FALSE}
library(kableExtra)
library(gridExtra)
library(dplyr)
library(MSnSet.utils)
library(ggplot2)
library(clusterProfiler)
library(amlresistancenetworks)
source("../util/synapseUtil.R")
source("../util/loadData.R")

## Using synapse ID to fecth table + metadata
dat <- load.DIA.data("syn26182854")

prot.dat <- dat[["Long form data"]]
meta <- dat[["Metadata"]] 
colnames(meta)[4] <- "Trametinib"
rownames(meta) <- meta$Sample

## Turning long form into matrix
## for any peptides with the same parent protein we average intensity.
prot.mat <- prot.dat %>%
  select(Protein, Sample, Intensity) %>%
  tidyr::pivot_wider(values_from='Intensity',names_from='Sample',
                     values_fn=list(Intensity = mean)) %>%
  tibble::column_to_rownames('Protein') %>%
  as.matrix()

## Creating MSnSet using the crosstab and metadata
m <- MSnSet(exprs = prot.mat, pData = meta)

```

Below we have the metadata for this experiment. We have a total of 34 samples
and 11 proteins after rolling up from peptide level data. 
The samples are split in two ways: there are `Pre` and `Post` treatment
groups, as well as 3 cell types, `CD14`, `CD34` and `MNC`. In this initial analysis we
compare the `Pre` and `Post` groups within each cell type and across cell types, we also
compare each pair of cell types to each other.

The targeted data has intensities present for all 34 samples. We have rolled up from the 
peptide level to the protein level

```{r pressure, echo=FALSE}
show <- meta %>%
  select(-Sample)
kbl(show) %>%
  kable_paper("hover", full_width = F)

```

$$\\[.2in]$$

# PCA plots

Below we collect two PCA plots for this data: by treatment, and by cell type.

```{r PCA by treatment}
p <- plot_pca_v4(m, "Trametinib", show.ellipse = F) +
  ggtitle("PCA by Treatment") + 
  theme(plot.title = element_text(size = 13, hjust = 0.5))

p

```



```{r PCA by cell type}
p <- plot_pca_v4(m, "CellType", show.ellipse = F) +
  ggtitle("PCA by Cell Type") + 
  theme(plot.title = element_text(size = 13, hjust = 0.5))

p

```

$$\\[.2in]$$

# Pre-Trametinib vs Post-Trametinib

Next, we compare the Pre and Post treatment groups within each cell type individually.


```{r run differential expression, include=FALSE}

diff.results <- list()

for (type in c("CD14", "CD34", "MNC")) {
  xx <- prot.mat[, meta$CellType == type]
  
  pre.cols <- colnames(xx)[grepl("Pre", colnames(xx))]
  post.cols <- colnames(xx)[grepl("Post", colnames(xx))]

  limma <- limmaTwoFactorDEAnalysis(xx, pre.cols, post.cols)
  
  diff.results[[type]] <- limma
}

```

$$\\[.1in]$$

### CD14 Cell Type

Restricting to those `CD14` cell types, we compare the pre and post-treatment group using differential
expression.


```{r CD14 differential expression show}
show <- diff.results[["CD14"]] %>%
  select(-featureID, -t, -B, -AveExpr) %>% head(.)
kbl(show) %>%
  kable_paper("hover", full_width = F)

```


$$\\[.1in]$$

### CD34 Cell Type

Restricting to those `CD34` cell types, we compare the pre and post-treatment group using differential
expression.


```{r CD34 differential expression show}
show <- diff.results[["CD34"]] %>%
  select(-featureID, -t, -B, -AveExpr) %>% head(.)
kbl(show) %>%
  kable_paper("hover", full_width = F)

```

$$\\[.1in]$$

### MNC Cell Type

Restricting to those `MNC` cell types, we compare the pre and post-treatment group using differential
expression.


```{r MNC differential expression show}
show <- diff.results[["MNC"]] %>%
  select(-featureID, -t, -B, -AveExpr) %>% head(.)
kbl(show) %>%
  kable_paper("hover", full_width = F)

```

$$\\[.1in]$$

### Pre vs Post (All cell types)


```{r Pre vs Post (all cells), include=FALSE}

xx <- prot.mat

pre.cols <- colnames(xx)[grepl("Pre", colnames(xx))]
post.cols <- colnames(xx)[grepl("Post", colnames(xx))]

limma <- limmaTwoFactorDEAnalysis(xx, pre.cols, post.cols) 

```

We also compare the two treatment groups without restricting the cell type.=

```{r all cells differential expression show}
show <- limma %>%
  select(-featureID, -t, -B, -AveExpr) %>% head(.)
kbl(show) %>%
  kable_paper("hover", full_width = F)

```


$$\\[.2in]$$

# Comparing Cell Types

Next, we compare all the cell types to each other, pair by pair. 
We again use differential expression and GSEA with the KEGG database in order 
to compare the groups.

```{r cell type comparisons, include=FALSE}
limma.results <- list()

for (type in c("CD14", "CD34", "MNC")) {
  types <- meta$CellType != type
  xx <- prot.mat[, types]
  
  types <- unique(meta$CellType[types])
  group1 <- colnames(xx)[grepl(types[1], colnames(xx))]
  group2 <- colnames(xx)[grepl(types[2], colnames(xx))]
  
  limma <- limmaTwoFactorDEAnalysis(xx, group1, group2)
  limma.results[[type]] <- limma
}

```

$$\\[.1in]$$

### CD14 vs CD34

We remove the `MNC` samples, then compare the `CD14` and `CD34` cell types using 
differential expression.


```{r cell types MNC differential expression show}
show <- limma.results[["MNC"]] %>%
  select(-featureID, -t, -B, -AveExpr) %>% head(.)
kbl(show) %>%
  kable_paper("hover", full_width = F)

```

$$\\[.1in]$$

### CD34 vs MNC

We remove the `CD14` samples, then compare the `CD34` and `MNC` cell types using 
differential expression.


```{r cell types CD14 differential expression show}
show <- limma.results[["CD14"]] %>%
  select(-featureID, -t, -B, -AveExpr) %>% head(.)
kbl(show) %>%
  kable_paper("hover", full_width = F)

```

$$\\[.1in]$$

### CD14 vs MNC

We remove the `CD34` samples, then compare the `CD14` and `MNC` cell types using 
differential expression.


```{r cell types CD34 differential expression show}
show <- limma.results[["CD34"]] %>%
  select(-featureID, -t, -B, -AveExpr) %>% head(.)
kbl(show) %>%
  kable_paper("hover", full_width = F)

```


# Plotting the proteins

Finally, we make boxplots of the 11 proteins, grouped by treatment and cell type.

```{r boxplotting intensities, include=FALSE}
proteins <- rownames(prot.mat)

plots <- list()

for (protein in proteins){
  xx <- data.frame(Intensity = prot.mat[protein, ], Sample = colnames(prot.mat)) %>%
    left_join(meta, by = "Sample") %>%
    mutate(Trametinib = factor(Trametinib, levels = c("Pre", "Post")))
  
  p <- ggplot(xx, aes(y = Intensity, fill = CellType, x = Trametinib)) + geom_boxplot() +
    ggtitle(protein) + theme(plot.title = element_text(hjust = 0.5, size = 10),
                             axis.title.x = element_text(size = 8),
                             axis.title.y = element_text(size = 8))
  
  plots[[protein]] <- p
}

```


```{r show plots 1}
layout <- rbind(c(1,1,2,2),
                c(3,3,4,4))

grid.arrange(grobs = plots[1:4], layout_matrix = layout)

```


```{r show plots 2}
layout <- rbind(c(5,5,6,6),
                c(7,7,8,8))

grid.arrange(grobs = plots[5:8], layout_matrix = layout)

```


```{r show plots 3}
layout <- rbind(c(9,9,10,10),
                c(NA,11,11,NA))

grid.arrange(grobs = plots[9:11], layout_matrix = layout)

```












