---
title: "CD14 Cell Line"
author: "Camilo Posso"
date: "09/27/2021"
output: 
  html_document:
    code_folding: hide
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

```{r libraries, include=FALSE}
library(kableExtra)
library(gridExtra)
library(dplyr)
library(MSnSet.utils)
library(ggplot2)
library(clusterProfiler)
library(amlresistancenetworks)
source("../util/synapseUtil.R")
source("../util/loadData.R")

## Using synapse ID to fecth table + metadata
dat <- load.DIA.data("syn26164985")

prot.dat <- dat[["Long form data"]]
meta <- dat[["Metadata"]] 
colnames(meta)[4] <- "Trametinib"
rownames(meta) <- meta$Sample

## Turning long form into matrix
prot.mat <- prot.dat %>%
  select(Protein, Sample, Intensity) %>%
  tidyr::pivot_wider(values_from='Intensity',names_from='Sample',
                     values_fn=list(Intensity = mean)) %>%
  tibble::column_to_rownames('Protein') %>%
  as.matrix()

```


```{r subset to cell line and define params, include=FALSE}
cell.line <- "CD14"
meta <- meta[meta$CellType == cell.line, ]

prot.mat <- prot.mat[, grepl(cell.line, colnames(prot.mat))]
n.samples <- ncol(prot.mat)

## Filter out fully NA rows
idx <- rowSums(is.na(prot.mat)) == ncol(prot.mat)
prot.mat <- prot.mat[!idx, ]

## 10 hand picked colors, chosen to be distinct from each other (normal color vision)
colors.chosen <- c("darkorange2", "dodgerblue2", "firebrick3",
                   "lawngreen", "gray69", "darkorchid1", 
                   "forestgreen", "navy", "deeppink2",
                   "lightsteelblue2", "goldenrod1")
names(colors.chosen) <- colnames(prot.mat)

```



```{r restrict samples}
restrict <- F

if (restrict) {
  samples <- sort(colSums(is.na(prot.mat)))[1:6] %>%
    names()
} else {
  samples <- sort(colSums(is.na(prot.mat))) %>%
    names()
}

prot.mat <- prot.mat[, samples]

colors.use <- colors.chosen[samples]

meta <- meta[rownames(meta) %in% samples, ]

## Filter out fully NA rows
idx <- rowSums(is.na(prot.mat)) == ncol(prot.mat)
prot.mat <- prot.mat[!idx, ]

## Creating MSnSet using the crosstab and metadata
m <- MSnSet(exprs = prot.mat, pData = meta[colnames(prot.mat), ])

```


```{r functions, include=FALSE}

filter.DreamAI <- function(mat, prop.NA, step) {

  # remove rows where all data is NA
  print(nrow(mat))
  mat <- mat[rowSums(is.na(mat)) != ncol(mat), ]
  print(nrow(mat))
  
  # Any features with more than 90% missing data are thrown out. Needed for using
  # DreamAI!! using a slightly stricter filter in order to obtain under 90% missing
  # in the columns as well!!
  N <- ncol(mat)
  keeping <- rowSums(is.na(mat)) < prop.NA*N 
  mat <- mat[keeping, ]
  
  ## Even with the filtering above, we still have a few columns with 93% or so 
  ## missing data so we have to drop a few more rows, strategically.

  while(any(colSums(is.na(mat)) >= 0.9*nrow(mat))){
    
  ## We weigh each intensity from a particular sample according to how much data
  ## that sample is missing. Then we aggregate these into a feature score.
  ## For a given feature, the higher the score, the more intensities from highly 
  ## missing samples are present in that feature.
  weights <- colSums(is.na(mat))/(0.9*nrow(mat))
  weights <- matrix(weights, nrow = nrow(mat), ncol = 30, byrow = TRUE)
  
  score <- as.numeric(!is.na(mat))*weights
  rownames(score) <- rownames(mat)
  importance <- rowSums(score) %>%
    sort(decreasing = F) %>%
    names()
  
  keeping <- importance[step:length(importance)]
  mat <- mat[keeping, ]
  }
  
  return(mat)
}


make.NA <- function(mat, pseudo.NA) {
  
  present <- which(!is.na(mat))
  total <- nrow(mat)*ncol(mat)
  
  N <- length(present)*pseudo.NA
  pseudo.missing <- sample(present, N)
  
  mat[pseudo.missing] <- NA
  missing.index <- arrayInd(pseudo.missing, c(nrow(mat), ncol(mat)))
  missing.index[, 1] <- rownames(mat)[missing.index[, 1]]
  missing.index[, 2] <- colnames(mat)[as.numeric(missing.index[, 2])]
  return(list("data" = mat, "pseudo NA index" = missing.index))
}


```


```{r prepare datasets}

params.NA <- c(0.2, 0.4, 0.6)
params.pseudoNA <- c(0.1, 0.1, 0.1)

```


Below we have the metadata for this experiment. We have a total of `r n.samples` samples
and `r nrow(prot.mat)` Proteins. In this initial analysis we compare the `Pre` and 
`Post` groups, and investigate whether the imputation of the missing values is meaningful.

```{r pressure, echo=FALSE}
show <- meta %>%
  select(-Sample)
kbl(show) %>%
  kable_paper("hover", full_width = F)

```

$$\\[.2in]$$

# Missing data

First, notice the large amount of missing values present. Below we summarize the distribution
of NA values in the data by counting the number of missing intensities per protein.


```{r missing data, include=FALSE}
library(scales)

missing <- rowSums(is.na(prot.mat))
dff.missing = data.frame(missing = missing)

breaks = seq(-0.25, ncol(prot.mat) + 0.25, 0.5)

p1 <- ggplot(dff.missing, aes(x = missing)) + geom_histogram(breaks = breaks) +
  ylab("") + xlab("Number missing") +
  ggtitle("Number missing per protein") +
  scale_x_continuous(breaks = pretty_breaks()) +
  theme(plot.title = element_text(hjust = 0.5))

dff.sample.missing <- data.frame(missing = colSums(is.na(prot.mat)),
                                 Sample = colnames(prot.mat))
  
p2 <- ggplot(dff.sample.missing, aes(x = Sample, fill = Sample, y = missing)) +
  geom_bar(stat = "identity") +
  ggtitle("Missing per sample") + 
  theme(plot.title = element_text(hjust = 0.5, size = 13),
        axis.text.x = element_text(vjust = 0.5, angle = 90)) +
  scale_fill_manual(values = colors.use) 

complete.features <- length(which(rowSums(is.na(prot.mat)) == 0))


```



```{r}
p1

```


```{r}
p2

```


$$\\[.2in]$$

# PCA plots

Note that only the `r complete.features` proteins with no missing data can be 
used when making a PCA plot.

```{r PCA by treatment}
p <- plot_pca_v4(m, "Trametinib", show.ellipse = F) +
  ggtitle("PCA by Treatment") + 
  theme(plot.title = element_text(size = 13, hjust = 0.5))

p

```



$$\\[.2in]$$

# Pre-Trametinib vs Post-Trametinib

```{r include=FALSE}
missing.fraction <- 0.6

```

Next, we compare the Pre and Post treatment groups within each cell type individually.
Due to the large amount of missing data, we first restrict the samples by cell type, then 
filter the proteins such that at most `r missing.fraction*100`% of the data is missing 
within that cell type.

```{r run GSEA comparisons by cell, include=FALSE}

xx <- prot.mat
chosen.features <- xx %>%
  is.na() %>%
  rowSums()
chosen.features <- which(chosen.features <= ncol(xx)*missing.fraction) %>%
  names()

xx <- xx[chosen.features, ]
n.features <- length(chosen.features)

pre.cols <- colnames(xx)[grepl("Pre", colnames(xx))]
pre.means <- apply(xx[, pre.cols], 1, mean, na.rm = T)
post.cols <- colnames(xx)[grepl("Post", colnames(xx))]
post.means <- apply(xx[, post.cols], 1, mean, na.rm = T)

log.fold.change <- post.means - pre.means
log.fold.change <- sort(log.fold.change, decreasing = T)

set.seed(117)
rank.GSEA <- gseKEGG(geneList = log.fold.change,
               organism     = 'hsa',
               minGSSize    = 10,
               pvalueCutoff = 1,
               verbose      = T,
               keyType = 'uniprot')


```


```{r run differential expression, include=FALSE}

xx <- prot.mat
chosen.features <- xx %>%
  is.na() %>%
  rowSums()
chosen.features <- which(chosen.features <= ncol(xx)*missing.fraction) %>%
  names()

xx <- xx[chosen.features, ]

pre.cols <- colnames(xx)[grepl("Pre", colnames(xx))]
post.cols <- colnames(xx)[grepl("Post", colnames(xx))]

limma <- limmaTwoFactorDEAnalysis(xx, pre.cols, post.cols)

```



# Imputation

Next, we try to test how far off imputed values are from their original
observed values. We do this by introducing `NEW` NA values to the dataset at random,
which we call pseudo NA values. We then compare the value imputed for the pseudo 
NA values to the original observed value. We do this a few times, filtering the 
original dataset at different levels each time.


```{r imputation test, include=FALSE}
library("cluster")
library("survival")
library("randomForest")
library("missForest")
library("glmnet")
library("Rcpp")
library("foreach")
library("itertools")
library("iterators")
library("Matrix")
library("impute")
library("DreamAI")

results <- list()
results.missing <- list()
results.count <- list()

set.seed(117)
for (i in 1:length(params.NA)) {
  param.NA <- params.NA[i]
  param.pNA <- params.pseudoNA[i]
  
  mat <- filter.DreamAI(prot.mat, param.NA, 3)
  
  mat.NA <- make.NA(mat, param.pNA)
  new.NA <- mat.NA$`pseudo NA index`
  mat.NA <- mat.NA$data
  mat.NA <- filter.DreamAI(mat.NA, 0.9, 3)

  mat.NA <- DreamAI(mat.NA,k=10,maxiter_MF = 10, ntree = 100,maxnodes = NULL,
                    maxiter_ADMIN=30, tol=10^(-2),gamma_ADMIN=0,gamma=50,
                    CV=FALSE,fillmethod="row_mean",maxiter_RegImpute=10,
                    conv_nrmse = 1e-6,iter_SpectroFM=40, out="Ensemble",
                    method = c("KNN", "ADMIN", "Brinn", 
                               "SpectroFM", "RegImpute"))[["Ensemble"]]
  
  new.NA <- new.NA[new.NA[, 1] %in% rownames(mat.NA), ]
  impute.diff <- mat.NA[new.NA[, 1], ] - mat[new.NA[, 1], ]
  impute.diff <- sapply(1:nrow(impute.diff), function(x) {
    impute.diff[x, new.NA[x, 2]]
  })
  impute.diff <- data.frame(Difference = impute.diff,
                            Sample = new.NA[, 2]) %>%
    filter(!is.na(Difference))

  name <- paste("param NA", param.NA, "param pNA", param.pNA)
  print(name)
  results[[name]] <- impute.diff
  
  dff.missing <- colSums(is.na(mat))
  dff.missing <- data.frame(Sample = names(dff.missing), Missing = dff.missing)
  results.missing[[name]] <- dff.missing
  results.count[[name]] <- nrow(mat.NA)
  
}

```



### Differential Expression & GSEA


Using these proteins, we compare the pre and post-treatment group using differential
expression and rank based GSEA with the KEGG database.


```{r CD14 differential expression show}
show <- limma %>%
  select(-featureID, -t, -B, -AveExpr) %>% head(.)
kbl(show) %>%
  kable_paper("hover", full_width = F)

```

As no pathway showed significant enrichment, we show the most significantly
enriched pathways instead.

```{r Rank Based GSEA CD14 show}
show <- rank.GSEA %>%
  select(Description, setSize, NES, pvalue, p.adjust) %>%
  head(10)

kbl(show) %>%
  kable_paper("hover", full_width = T)
```


### At most `r params.NA[1]*100`% missing per protein, `r params.pseudoNA[1]*100`% pseudo NA introduced.


```{r show differences 1, include=FALSE}
name <- paste("param NA", params.NA[1], "param pNA", params.pseudoNA[1])

dff <- results[[name]]
dff.missing <- results.missing[[name]]

p1 <- ggplot(dff, aes(x = Difference, fill = Sample)) + geom_histogram(breaks = seq(-5.25,5.25,0.5)) +
  ggtitle("Difference between original & pseudo NA imputed") +
  theme(plot.title = element_text(hjust = 0.5, size = 13)) +
  scale_fill_manual(values = colors.use)

p2 <- ggplot(dff.missing, aes(x = Sample, y = Missing, fill = Sample)) + geom_bar(stat = "identity") +
  ggtitle("Missing per sample") + 
  theme(plot.title = element_text(hjust = 0.5, size = 13),
        axis.text.x = element_text(vjust = 0.5, angle = 90)) +
  scale_fill_manual(values = colors.use) 

MSE <- sqrt(sum((dff$Difference)^2)/nrow(dff))

```



We filter the original data to have at most `r params.NA[1]*100`% of its entries per protein
as NA values. We then randomly choose `r params.pseudoNA[1]*100`% of the
present intensities and replace them with NA values. Our imputed dataset
has `r results.count[1]` proteins in total. Below we see the distribution 
of the differences between the imputed (pseudo NA) values and the original values. 

The mean squared error between the imputed pseudo NA values and the original 
values is `r MSE`.


```{r}
p1

```

```{r}
p2

```



### At most `r params.NA[2]*100`% missing per protein, `r params.pseudoNA[2]*100`% pseudo NA introduced.



```{r show differences 2, include=FALSE}
name <- paste("param NA", params.NA[2], "param pNA", params.pseudoNA[2])

dff <- results[[name]]
dff.missing <- results.missing[[name]]

p1 <- ggplot(dff, aes(x = Difference, fill = Sample)) + geom_histogram(breaks = seq(-5.25,5.25,0.5)) +
  ggtitle("Difference between original & pseudo NA imputed") +
  theme(plot.title = element_text(hjust = 0.5, size = 13)) +
  scale_fill_manual(values = colors.use)

p2 <- ggplot(dff.missing, aes(x = Sample, y = Missing, fill = Sample)) + geom_bar(stat = "identity") +
  ggtitle("Missing per sample") + 
  theme(plot.title = element_text(hjust = 0.5, size = 13),
        axis.text.x = element_text(vjust = 0.5, angle = 90)) +
  scale_fill_manual(values = colors.use) 

MSE <- sqrt(sum((dff$Difference)^2)/nrow(dff))

```



We filter the original data to have at most `r params.NA[2]*100`% of its entries per protein
as NA values. We then randomly choose `r params.pseudoNA[2]*100`% of the
present intensities and replace them with NA values.  Our imputed dataset
has `r results.count[2]` proteins in total. Below we see the distribution 
of the differences between the imputed (pseudo NA) values and the original values. 

The mean squared error between the imputed pseudo NA values and the original 
values is `r MSE`.


```{r}
p1

```

```{r}
p2

```

### At most `r params.NA[3]*100`% missing per protein, `r params.pseudoNA[3]*100`% pseudo NA introduced.



```{r show differences 3, include=FALSE}
name <- paste("param NA", params.NA[3], "param pNA", params.pseudoNA[3])

dff <- results[[name]]
dff.missing <- results.missing[[name]]

p1 <- ggplot(dff, aes(x = Difference, fill = Sample)) + geom_histogram(breaks = seq(-5.25,5.25,0.5)) +
  ggtitle("Difference between original & pseudo NA imputed") +
  theme(plot.title = element_text(hjust = 0.5, size = 13)) +
  scale_fill_manual(values = colors.use)

p2 <- ggplot(dff.missing, aes(x = Sample, y = Missing, fill = Sample)) + geom_bar(stat = "identity") +
  ggtitle("Missing per sample") + 
  theme(plot.title = element_text(hjust = 0.5, size = 13),
        axis.text.x = element_text(vjust = 0.5, angle = 90))  +
  scale_fill_manual(values = colors.use)

MSE <- sqrt(sum((dff$Difference)^2)/nrow(dff))

```



We filter the original data to have at most `r params.NA[3]*100`% of its entries per protein
as NA values. We then randomly choose `r params.pseudoNA[3]*100`% of the
present intensities and replace them with NA values.  Our imputed dataset
has `r results.count[3]` proteins in total. Below we see the distribution 
of the differences between the imputed (pseudo NA) values and the original values. 


The mean squared error between the imputed pseudo NA values and the original 
values is `r MSE`.


```{r}
p1

```

```{r}
p2

```


