---
title: "Sunil Ex17 Global differential expression & GSEA"
author: "Camilo Posso"
date: "09/13/2021"
output: 
  html_document:
    code_folding: hide
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

```{r libraries, include=FALSE}
library(kableExtra)
library(gridExtra)
library(dplyr)
library(MSnSet.utils)
library(ggplot2)
library(clusterProfiler)
library(amlresistancenetworks)
source("util/synapseUtil.R")
source("util/loadData.R")

## Using synapse ID to fecth table + metadata
dat <- load.DIA.data("syn26164985")

prot.dat <- dat[["Long form data"]]
meta <- dat[["Metadata"]] 
colnames(meta)[4] <- "Trametinib"
rownames(meta) <- meta$Sample

## Turning long form into matrix
prot.mat <- prot.dat %>%
  select(Protein, Sample, Intensity) %>%
  tidyr::pivot_wider(values_from='Intensity',names_from='Sample',
                     values_fn=list(LogRatio=mean)) %>%
  tibble::column_to_rownames('Protein') %>%
  as.matrix()

## Creating MSnSet using the crosstab and metadata
m <- MSnSet(exprs = prot.mat, pData = meta)

```

Below we have the metadata for this experiment. We have a total of 30 samples
and 2319 proteins. The samples are split in two ways: there are `Pre` and `Post` treatment
groups, as well as 3 cell types, `CD14`, `CD34` and `MNC`. In this initial analysis we
compare the `Pre` and `Post` groups within each cell type and across cell types, we also
compare each pair of cell types to each other.

```{r pressure, echo=FALSE}
show <- meta %>%
  select(-Sample)
kbl(show) %>%
  kable_paper("hover", full_width = F)

```

$$\\[.2in]$$

# Missing data

First, notice the large amount of missing values present. Below we summarize the distribution
of NA values in the data by counting the number of missing intensities per feature.

```{r missing data}
missing <- rowSums(is.na(prot.mat))
dff = data.frame(missing = missing)

breaks = seq(-0.25, ncol(prot.mat) + 0.25, 0.5)

p1 <- ggplot(dff, aes(x = missing)) + geom_histogram(breaks = breaks) +
  ylab("") + xlab("Number missing") +
  ggtitle("Number missing per feature") +
  theme(plot.title = element_text(hjust = 0.5))
p1

```

We have 2 treatment groups `Pre` and `Post`, as well as 3 cell types
`CD14`, `CD34` and `MNC`. Below we further restrict to these groups when counting the number
of missing values per feature.

```{r missing data per group, include=FALSE}
groups <- list(
  "Pre" = meta$Trametinib == "Pre",
  "Post" = meta$Trametinib == "Post",
  "CD14" = meta$CellType == "CD14",
  "CD34" = meta$CellType == "CD34",
  "MNC" = meta$CellType == "MNC")

plots <- list()

for (name in names(groups)) {
  filter <- groups[[name]]
  xx <- prot.mat[, filter]
  missing <- rowSums(is.na(xx))
  dff <- data.frame(missing = missing)
  breaks <- seq(-0.25, ncol(xx) + 0.25, 0.5)
  plot.title <- paste("Missing per feature in", name, "group")
  
  p <- ggplot(dff, aes(x = missing)) + geom_histogram(breaks = breaks) +
    ylab("") + xlab("Number missing") +
    ggtitle(plot.title) +
    theme(plot.title = element_text(hjust = 0.5, size = 8.5),
          axis.title.x = element_text(size = 7.5))
  
  plots <- append(plots, list(p))
}
```



```{r show missing plots}
layout <- rbind(c(1,1,1,2,2,2),
                c(3,3,4,4,5,5))
grid.arrange(grobs = plots, layout_matrix = layout)

```

$$\\[.2in]$$

# PCA plots

Below we collect two PCA plots for this data,

```{r PCA by treatment}
p <- plot_pca_v4(m, "Trametinib", show.ellipse = F) +
  ggtitle("PCA by Treatment") + 
  theme(plot.title = element_text(size = 13, hjust = 0.5))

p

```



```{r PCA by cell type}
p <- plot_pca_v4(m, "CellType", show.ellipse = F) +
  ggtitle("PCA by Cell Type") + 
  theme(plot.title = element_text(size = 13, hjust = 0.5))

p

```

$$\\[.2in]$$

# Pre-Trametinib vs Post-Trametinib

```{r include=FALSE}
missing.fraction <- 0.6

```

Next, we compare the Pre and Post treatment groups within each cell type individually.
Due to the large amount of missing data, we first restrict the samples by cell type, then 
filter the features such that at most `r missing.fraction*100`% of the data is missing 
within that cell type.

```{r run GSEA comparisons by cell, include=FALSE}

gsea.results <- list()

for (type in c("CD14", "CD34", "MNC")) {
  xx <- prot.mat[, meta$CellType == type]
  chosen.features <- xx %>%
    is.na() %>%
    rowSums()
  chosen.features <- which(chosen.features <= ncol(xx)*missing.fraction) %>%
    names()
  
  xx <- xx[chosen.features, ]
  
  pre.cols <- colnames(xx)[grepl("Pre", colnames(xx))]
  pre.means <- apply(xx[, pre.cols], 1, mean, na.rm = T)
  post.cols <- colnames(xx)[grepl("Post", colnames(xx))]
  post.means <- apply(xx[, post.cols], 1, mean, na.rm = T)
  
  log.fold.change <- post.means - pre.means
  log.fold.change <- sort(log.fold.change, decreasing = T)
  
  set.seed(117)
  rank.GSEA <- gseKEGG(geneList = log.fold.change,
                 organism     = 'hsa',
                 minGSSize    = 10,
                 pvalueCutoff = 1,
                 verbose      = T,
                 keyType = 'uniprot')
  gsea.results[[type]] <- rank.GSEA
}

```


```{r run differential expression, include=FALSE}

diff.results <- list()

for (type in c("CD14", "CD34", "MNC")) {
  xx <- prot.mat[, meta$CellType == type]
  chosen.features <- xx %>%
    is.na() %>%
    rowSums()
  chosen.features <- which(chosen.features <= ncol(xx)*missing.fraction) %>%
    names()
  
  xx <- xx[chosen.features, ]
  
  pre.cols <- colnames(xx)[grepl("Pre", colnames(xx))]

  post.cols <- colnames(xx)[grepl("Post", colnames(xx))]

  limma <- limmaTwoFactorDEAnalysis(xx, pre.cols, post.cols) %>%
  rename(value = "logFC")
  
  diff.results[[type]] <- limma
}

```

$$\\[.1in]$$

### CD14 Cell Type

Restricting to those `CD14` cell types, we then filter out all features with more than
`r missing.fraction`% of the data missing within the `CD14` group. 

Using these features, we compare the pre and post-treatment group using differential
expression and rank based GSEA with the KEGG database.


```{r CD14 differential expression show}
show <- diff.results[["CD14"]] %>%
  select(-featureID, -t, -B, -AveExpr) %>% head(.)
kbl(show) %>%
  kable_paper("hover", full_width = F)

```

As no pathway showed significant enrichment, we show the most significantly
enriched pathways instead.

```{r Rank Based GSEA CD14 show}
show <- gsea.results[["CD14"]] %>%
  select(Description, setSize, NES, pvalue, p.adjust) %>%
  head(10)

kbl(show) %>%
  kable_paper("hover", full_width = T)
```

$$\\[.1in]$$

## CD34 Cell Type

Restricting to those `CD34` cell types, we then filter out all features with more than
`r missing.fraction`% of the data missing within the `CD34` group. 

Using these features, we compare the pre and post-treatment group using differential
expression and rank based GSEA with the KEGG database.


```{r CD34 differential expression show}
show <- diff.results[["CD34"]] %>%
  select(-featureID, -t, -B, -AveExpr) %>% head(.)
kbl(show) %>%
  kable_paper("hover", full_width = F)

```

As no pathway showed significant enrichment, we show the most significantly
enriched pathways instead.

```{r Rank Based GSEA CD34 show}
show <- gsea.results[["CD34"]] %>%
  select(Description, setSize, NES, pvalue, p.adjust) %>%
  head(10)

kbl(show) %>%
  kable_paper("hover", full_width = T)
```

$$\\[.1in]$$

## MNC Cell Type

Restricting to those `MNC` cell types, we then filter out all features with more than
`r missing.fraction`% of the data missing within the `MNC` group. 

Using these features, we compare the pre and post-treatment group using differential
expression and rank based GSEA with the KEGG database.


```{r MNC differential expression show}
show <- diff.results[["MNC"]] %>%
  select(-featureID, -t, -B, -AveExpr) %>% head(.)
kbl(show) %>%
  kable_paper("hover", full_width = F)

```

As no pathway showed significant enrichment, we show the most significantly
enriched pathways instead.

```{r Rank Based GSEA MNC show}
show <- gsea.results[["MNC"]] %>%
  select(Description, setSize, NES, pvalue, p.adjust) %>%
  head(10)

kbl(show) %>%
  kable_paper("hover", full_width = T)
```

$$\\[.1in]$$

## Pre vs Post (All cell types)

We also compare the two treatment groups without restricting the cell type. We still
filter out all features with more than `r missing.fraction`% of the data missing.

```{r Pre vs Post (all cells), include=FALSE}

xx <- prot.mat
chosen.features <- xx %>%
  is.na() %>%
  rowSums()
chosen.features <- which(chosen.features <= ncol(xx)*missing.fraction) %>%
  names()

xx <- xx[chosen.features, ]

pre.cols <- colnames(xx)[grepl("Pre", colnames(xx))]
pre.means <- apply(xx[, pre.cols], 1, mean, na.rm = T)
post.cols <- colnames(xx)[grepl("Post", colnames(xx))]
post.means <- apply(xx[, post.cols], 1, mean, na.rm = T)

log.fold.change <- post.means - pre.means
log.fold.change <- sort(log.fold.change, decreasing = T)

set.seed(117)
rank.GSEA <- gseKEGG(geneList = log.fold.change,
               organism     = 'hsa',
               minGSSize    = 10,
               pvalueCutoff = 1,
               verbose      = T,
               keyType = 'uniprot')

limma <- limmaTwoFactorDEAnalysis(xx, pre.cols, post.cols) %>%
  rename(value = "logFC")


```


```{r all cells differential expression show}
show <- limma %>%
  select(-featureID, -t, -B, -AveExpr) %>% head(.)
kbl(show) %>%
  kable_paper("hover", full_width = F)

```

As no pathway showed significant enrichment, we show the most significantly
enriched pathways instead.

```{r Rank Based GSEA all cells show}
show <- rank.GSEA %>%
  select(Description, setSize, NES, pvalue, p.adjust) %>%
  head(10)

kbl(show) %>%
  kable_paper("hover", full_width = T)
```

$$\\[.2in]$$

# Comparing Cell Types

Finally, we compare all the cell types to each other, pair by pair. 
We again use differential expression and GSEA with the KEGG database in order 
to compare the groups.

```{r cell type comparisons, include=FALSE}
limma.results <- list()
gsea.results <- list()

for (type in c("CD14", "CD34", "MNC")) {
  types <- meta$CellType != type
  xx <- prot.mat[, types]
  
  chosen.features <- xx %>%
    is.na() %>%
    rowSums()
  chosen.features <- which(chosen.features <= ncol(xx)*missing.fraction) %>%
    names()
  xx <- xx[chosen.features, ]
  
  types <- unique(meta$CellType[types])
  group1 <- colnames(xx)[grepl(types[1], colnames(xx))]
  group1.means <- apply(xx[, group1], 1, mean, na.rm = T)
  group2 <- colnames(xx)[grepl(types[2], colnames(xx))]
  group2.means <- apply(xx[, group2], 1, mean, na.rm = T)
  
  limma <- limmaTwoFactorDEAnalysis(xx, group1, group2)
  limma.results[[type]] <- limma
  
  log.fold.change <- group2.means - group1.means
  log.fold.change <- sort(log.fold.change, decreasing = T)
  
  set.seed(117)
  rank.GSEA <- gseKEGG(geneList = log.fold.change,
                 organism     = 'hsa',
                 minGSSize    = 10,
                 pvalueCutoff = 1,
                 verbose      = T,
                 keyType = 'uniprot')
  gsea.results[[type]] <- rank.GSEA
}

```

$$\\[.1in]$$

## CD14 vs CD34

First, we remove the `MNC` samples, then we filter out all features with more than
`r missing.fraction`% of the data missing within. 

Using these features, we compare the `CD14` and `CD34` cell types using differential
expression and rank based GSEA with the KEGG database.


```{r cell types MNC differential expression show}
show <- limma.results[["MNC"]] %>%
  select(-featureID, -t, -B, -AveExpr) %>% head(.)
kbl(show) %>%
  kable_paper("hover", full_width = F)

```

As no pathway showed significant enrichment, we show the most significantly
enriched pathways instead. Positive NES scores indicate increased activity in the
`CD34` samples.

```{r Rank Based GSEA cell types MNC show}
show <- gsea.results[["MNC"]] %>%
  select(Description, setSize, NES, pvalue, p.adjust) %>%
  head(10)

kbl(show) %>%
  kable_paper("hover", full_width = T)
```

$$\\[.1in]$$

## CD34 vs MNC

First, we remove the `CD14` samples, then we filter out all features with more than
`r missing.fraction`% of the data missing within. 

Using these features, we compare the `CD34` and `MNC` cell types using differential
expression and rank based GSEA with the KEGG database.


```{r cell types CD14 differential expression show}
show <- limma.results[["CD14"]] %>%
  select(-featureID, -t, -B, -AveExpr) %>% head(.)
kbl(show) %>%
  kable_paper("hover", full_width = F)

```

As no pathway showed significant enrichment, we show the most significantly
enriched pathways instead. Positive NES scores indicate increased activity in the
`MNC` samples.

```{r Rank Based GSEA cell types CD14 show}
show <- gsea.results[["CD14"]] %>%
  select(Description, setSize, NES, pvalue, p.adjust) %>%
  head(10)

kbl(show) %>%
  kable_paper("hover", full_width = T)
```

$$\\[.1in]$$

## CD14 vs MNC

First, we remove the `CD34` samples, then we filter out all features with more than
`r missing.fraction`% of the data missing within. 

Using these features, we compare the `CD14` and `MNC` cell types using differential
expression and rank based GSEA with the KEGG database.


```{r cell types CD34 differential expression show}
show <- limma.results[["CD34"]] %>%
  select(-featureID, -t, -B, -AveExpr) %>% head(.)
kbl(show) %>%
  kable_paper("hover", full_width = F)

```

As no pathway showed significant enrichment, we show the most significantly
enriched pathways instead. Positive NES scores indicate increased activity in the
`MNC` samples.

```{r Rank Based GSEA cell types CD34 show}
show <- gsea.results[["CD34"]] %>%
  select(Description, setSize, NES, pvalue, p.adjust) %>%
  head(10)

kbl(show) %>%
  kable_paper("hover", full_width = T)
```










